DynaProg for Scala

### 2013-01-19 docs/report/benchmarks.tex
More report.

\subsection{Metrics}
The main metrix of interest is the running time. In an attempt to reduce the variance, we run 20 consecutive test and take the median running time (since median is less sensitive to outlier than the average).

As the device memory is quite limited, it seems interesting to also take into account the space usage. The space requirement limits the maximal size of addressable problems  on a particular hardware. This might be a concern for large problems, because they would require special adaptation to handle such cases both correctly and efficiently. However, simple solutions like using a device with larger memory or using main memory (if a $5\times$ slowdown is still acceptable) could solve this issue.

\subsection{Benchmarking platform}
Our benchmarking platform is a notebook with a Core i7-3720QM, 16Gb or main memory and an NVIDIA GeForce GT 650M running under MacOS X 10.8. Tests are made when the discrete GPU is not involved in rendering the user interface\footnote{This involves a workaround to keep the discrete GPU powered, while the CPU is involved in rendering graphics; source code available on demand.}. Due to impossibility to disable the watchdog timer in MacOS, CUDA kernels are limited to few seconds of running time before they are automatically aborted.

% ------------------------------------------------------------------------------------------------
{\center\color{red} \noindent\rule{16cm}{0.4pt} \\ XXX: CONTINUE HERE :XXX \\}
% ------------------------------------------------------------------------------------------------

{\color{red}\ol
\item Make LibRNA work with CUDA generated parsers
\item Finish the LibRNA section, write some conclusion ideas
\item Run benchmark and add results in report (automate, generate raw Matlab/TeX output)
\item Stabilize times, statistics, use CUDA profiler(?)
\ole}

\subsection{Matrix chain multiplication}
\subsubsection{CPU: DynaProg(Scala), hand optimized C, GAPC and ADPFusion}
XXX

@@

\subsection{Scala parsers}
{\color{red}
MatrixMult-512, Mac+JDK7\ul
\item Original: 24.937 sec / 24.658 sec
\item Optimized concatenation: 19.329 sec / 18.924 sec
\item Tabulation as arrays+inline: 14.976 sec / 14.849 sec
\ule}

>>>>

\subsection{Metrics}
The main metrics of interest is the running time. In an attempt to reduce the variance, we would like to run multiple consecutive test and take the median running time, since the median is less sensitive to outlier than the average\cite{perfeval}. Unfortunately, several factors hampers these ideal conditions. First the variance in the running time of CUDA kernels might be significant, in particular for short running time. This is due to the fact that the GPU needs to be 'warmed-up' before actual computation can happen. Similarly, the JVM is also subject to running time variance that is mainly due to the garbage collection\footnote{\url{http://www.oracle.com/technetwork/java/javase/gc-tuning-6-140523.html\#cms.overhead}} and JIT optimizations \cite{java_jit}.

Also the input and problem might introduce variance. As example, we can consider two extreme cases: matrix chain multiplication and Zuker RNA folding, with a test environment of 100 random inputs (of length respectively 512 and 80) and a GPU warmup of 10 computations. In this settings, matrix chain multiplication computations are executed in a perfectly constant time (0.127 seconds), which mean that we sufficiently reduced the noise. Oppositely, the Zuker RNA folding running times appear much more scattered as presented below:

\begin{figure}[H]\begin{center}\includegraphics[width=16cm]{inc/var_zuker.pdf}\end{center}
\caption{Zuker folding running time (seconds). Quartiles: 0.152, 0.183 (median), 0.214}\label{fig:var_zuker}\end{figure}

Using the QQplot\footnote{Quantile-to-quantile plot, used to compare two distributions against each other.}, the distribution is heavily tailed (has more results towards the ends of the range) than a Gaussian distribution (fig.~\ref{fig:var_zuker} center) but fits better an uniform distribution (fig.~\ref{fig:var_zuker} right). If we run multiple time the program over the same input, we obtain the same behavior as with the matrix multiplication (strictly identical time); hence we can conclude that Zuker is an input sensitive problem whereas matrix chain multiplication is not. It follows that we need to be careful to test with exactly identical input set different implementations.

As the device memory is quite limited, it seems interesting to also take into account the space usage. The space requirement limits the maximal size of addressable problems  on a particular hardware. This might be a concern for large problems, because they would require special adaptation to handle such cases both correctly and efficiently. However, this metric heavily depends on the problem and simple solutions like using a device with larger memory or using main memory (if a $5\times$ slowdown is still acceptable) could solve this issue, hence we do not consider this metric hereafter (except as an upper bound on the dimension of the input).

\subsection{Benchmarking platform}
Our benchmarking platform is an Apple notebook with a Core i7-3720QM with 16Gb of main memory and an NVIDIA GeForce GT 650M running under MacOS X 10.8 and Oracle JDK 1.7.0-10. A workaround (see listing~\ref{cpu_workaround}) allows us to use the CPU to render the user interface while leaving the graphic card available to execute CUDA kernels. Unfortunately, due to impossibility to disable the watchdog timer in MacOS, CUDA kernels are limited to few seconds of running time before they are automatically aborted.

% ----------------------------------------------
% MatrixMult-512, Mac+JDK7\ul
% Original: 24.937 sec / 24.658 sec
% Optimized concatenation: 19.329 sec / 18.924 sec
% Tabulation as arrays+inline: 14.976 sec / 14.849 sec
\newpage
% ------------------------------------------------------------------------------------------------
{\center\color{red} \noindent\rule{16cm}{0.4pt} \\ XXX: CONTINUE HERE :XXX \\}
% ------------------------------------------------------------------------------------------------
\subsection{Matrix chain multiplication}

\subsubsection{CPU: DynaProg(Scala), hand optimized C, GAPC and ADPFusion}
XXX

\begin{table}[H]\begin{center}\begin{tabular}{rllll}\toprule
\bf Size	& \bf DynaProg & \bf Optimized C & \bf GAPC & \bf ADP Fusion \\ \midrule
128		&		&		&		& -		\\
256		&		&		&		& -		\\
512		&		&		&		& -		\\
1024	&		&		&		& -		\\
2048	&		&		&		& -		\\
4096	&		&		&		& -		\\
8192	&		&		&		& -		\\ \bottomrule
\end{tabular}
\end{center}
\caption{Matrix chain multiplication results (CPU)}
\end{table}

{\color{red}\ol
\item Write some conclusion ideas
\item Run benchmark and add results in report (automate, generate raw Matlab/TeX output)
\item Stabilize times, statistics, use CUDA profiler(?)
\ole}

@@

XXX: future work: infer automatically "always non-empty" property

\subsection{Scala parsers}

### 2013-01-19 docs/report/conclusion.tex
Bug with bounded sequences identified !
More report.

\begin{verbatim}

recap of the report

recap of the results

big picture: with actual compiler technology we
Embedded DSL vs external DSL
Larger picture how to go further

>>>>

\begin{verbatim}
Recap of the report:
- Goal: generalize solving dynamic programming on sequences for GPUs
- classification of the problems; in terms of recurrences = complicated, grammar = easier
- general technique to describe backtracking information
- implemented in Scala+CUDA
- additional analysis to generate efficient code (1 result, memory alignment, scheduling)

Recap of the results:
- XXX

Big picture (Larger picture how to go further):
- Code generation and actual compiler technology
- Embedded DSL vs external DSL
- Only a subset of the dynamic programming problems, easy to parallelize (similar memory accesses and scheduling pattern/progress)
  - others are iterative or sparse, different scheduling => multiple scheduling patterns to be discovered and implemented efficiently, possibly in LMS (if time permits)

### 2013-01-19 docs/report/implementation.tex
More report.

Leveraging the insights provided by \cite{gpu_atlp} and \cite{gpu_barrier}, we started with a basic implementation (where each CUDA thread processes one matrix line) with three additional optimizations:\ul
\item Memory accesses must be coalesced (memory accesses account for a significant part of the total running time, according to both manufacturer documentation and experiments \cite{perfeval})
\item Synchronization between threads can be done according to \cite{gpu_barrier}, additionally, we can slightly loosen the synchronization restrictions, as the paper describes a thread barrier whereas we only require a condition on previous thread progress (except for the parallelogram case, where we still require a barrier).
\item Computation progresses element-wise along the diagonal (maximizes the parallelism level)
\item Thread block size = warp size (32) to benefit from implicit synchronization within warps
\ule

@@

% ----------------------------------------------
\subsubsection{Results discussion}\ul
\item \textbf{User interface:} It has been put in evidence in \cite{perfeval} that using the GPU exclusively for CUDA or in combination with UI display (Mac OS) affects the performance (GeForce 330M). With the newer architecture, this difference has been reduced to less than 3.5\%, decoupled UI and CUDA   performing best. So we can safely ignore this issue.
\item \textbf{Blocks synchronization:}\ul
	\item Removing {\tt \_\_threadfence()} before the synchronization is not syntactically correct but results still remains valid, this confirms the observation made by \cite{gpu_barrier}. Speedup for matrix size of 1024 are 67ms (parallelogram) 100ms (triangle) 180ms (rectangle).
	\item In the parallelogram case, using all threads to monitor other blocks status instead of the first one only results in a 6.4x speedup (22.72$\to$3.52ms) for the parallelogram.
	\ule
\item \textbf{Multiple threads per matrix cell:} in the case of a triangular matrix, at each step, the number of cells to be computed (on the diagonal) decrease while the computation complexity increases (there is one more dependency). According to \cite{gpu_atlp}, the solution lies in adaptive thread mapping, using more than one thread to compute one matrix cell, depending on the complexity. However, in our setup (memory layout+algorithm+hardware), we did not find any improvement by doing so. We want to explore the reason for that: we pose as hypothesis that the bandwidth is the bottleneck of our setup and test it.\ul
\item

>>>>

Leveraging the insights provided by \cite{gpu_atlp} and \cite{gpu_barrier}, we started with a basic implementation (where each CUDA thread processes one matrix line) with three additional optimizations:\ul
\item Memory accesses must be coalesced (memory accesses account for a significant part of the total running time, according to both manufacturer documentation and experiments \cite{perfeval_gpu})
\item Synchronization between threads can be done according to \cite{gpu_barrier}, additionally, we can slightly loosen the synchronization restrictions, as the paper describes a thread barrier whereas we only require a condition on previous thread progress (except for the parallelogram case, where we still require a barrier).
\item Computation progresses element-wise along the diagonal (maximizes the parallelism level)
\item Thread block size = warp size (32) to benefit from implicit synchronization within warps
\ule

@@

% ----------------------------------------------
\subsubsection{Results discussion}\ul
\item \textbf{User interface:} It has been put in evidence in \cite{perfeval_gpu} that using the GPU exclusively for CUDA or in combination with UI display (Mac OS) affects the performance (GeForce 330M). With the newer architecture, this difference has been reduced to less than 3.5\%, decoupled UI and CUDA   performing best. So we can safely ignore this issue.
\item \textbf{Blocks synchronization:}\ul
	\item Removing {\tt \_\_threadfence()} before the synchronization is not syntactically correct but results still remains valid, this confirms the observation made by \cite{gpu_barrier}. Speedup for matrix size of 1024 are 67ms (parallelogram) 100ms (triangle) 180ms (rectangle).
	\item In the parallelogram case, using all threads to monitor other blocks status instead of the first one only results in a 6.4x speedup (22.72$\to$3.52ms) for the parallelogram.
	\ule
\item \textbf{Multiple threads per matrix cell:} in the case of a triangular matrix, at each step, the number of cells to be computed (on the diagonal) decrease while the computation complexity increases (there is one more dependency). According to \cite{gpu_atlp}, the solution lies in adaptive thread mapping, using more than one thread to compute one matrix cell, depending on the complexity. However, in our setup (memory layout+algorithm+hardware), we did not find any improvement by doing so. We want to explore the reason for that: we pose as hypothesis that the bandwidth is the bottleneck of our setup and test it.\ul
\item
